---
title: 'snRNA-seq workflow: bioconductor'
author: "Saskia Freytag"
date: "08/04/2019"
output:
  html_document:
    toc: true
    theme: simplex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
```

## Prerequisities 

### Required libraries

```{r libraries}
library(DropletUtils)
library(scater)
library(scran)
library(umap)
```

```{r setup-ggplot2}
library(ggplot2)
theme_set(theme_classic())
```

## Import data

### Loading data

Unfortunately, Cell Ranger recently changed its naming convention. The file that
used to be called `genes.tsv` is now named `features.tsv`. The easiest 
work-around, until a change to `DropUtils`, is to rename the files.

```{r import}
samples <- list.files(here::here("data/raw"))
dat <- lapply(samples, function(x) 
  read10xCounts(here::here("data/raw", x, "mm10")))
names(dat) <- samples
```

## Filtering

### Filtering of nuclei

First, we filter any nuclei with greater than 10\% total mitochondrial 
contamination. 

```{r filter-mito}
for  (i in 1:length(dat)) {
  
  rowData(dat[[i]])$Mito <-  grepl("^mt-", rowData(dat[[i]])$Symbol)
  dat[[i]] <- calculateQCMetrics(dat[[i]], 
    feature_controls=list(Mt=rowData(dat[[i]])$Mito))

}

no_nuclei_mito <- sapply(dat, function(x) sum(colData(x)$pct_counts_Mt>10))

paste0("Nuclei with more than 5% mitochondrial percentage in ", samples, ": ",
      no_nuclei_mito, " (", round(no_nuclei_mito/sapply(dat, 
        function(x) dim(x)[2]),2)*100, "%)")

dat <- lapply(dat, function(x) x[,!colData(x)$pct_counts_Mt>10])
```

Next we filter any nuclei that have library sizes three standard below the 
median. We also filter any nuclei with a number of features three standard
deaviations below the median.

```{r filter-low}
libsize_drop <- lapply(dat, function(x) 
  isOutlier(x$total_counts, nmads=3, type="lower", log=TRUE))
feature_drop <- lapply(dat, function(x) 
  isOutlier(x$total_features_by_counts, nmads=3, type="lower", log=TRUE))

no_nuclei_libsize <- sapply(libsize_drop, function(x) sum(x))
no_nuclei_feature <- sapply(feature_drop, function(x) sum(x))

paste0("Nuclei not passing library size filter ", samples, ": ",
      no_nuclei_libsize, " (", round(no_nuclei_libsize/sapply(dat, function(x) dim(x)[2]),2)*100, "%)")

paste0("Nuclei not passing feature filter ", samples, ": ",
      no_nuclei_feature, " (", round(no_nuclei_feature/sapply(dat, function(x) dim(x)[2]),2)*100, "%)")

for (i in 1:length(samples)){
  
  dat[[i]] <- dat[[i]][, !(libsize_drop[[i]]|feature_drop[[i]])]

}
```

### Filtering of genes

## Filter mitochondrial genes

```{r filter-mito-genes}
dat <- lapply(dat, function(x) x[!rowData(x)$Mito,])
```

## Filter genes by expression

In order to filter genes we need to combine the data. Then we filter any gene
that is seen in less than 1\% of the data.

```{r filter-genes}
sce_counts <- Reduce(cbind, lapply(dat, function(x) counts(x)))

keep_genes <- apply(sce_counts, 1, function(x) 
  sum(x>0)>(dim(sce_counts)[2]*0.01))

dat <- lapply(dat, function(x) x[keep_genes,])
sce_counts <- sce_counts[keep_genes,]
```

## Normalization

```{r normalization}
#dat <- do.call(multiBatchNorm, dat)
dat <- lapply(dat, normalize)
```

### Batch correction

Perfom mnn correction to generate a common PCA representation for the samples.

```{r batch-correction}
dat_pca <- do.call(fastMNN, 
  c(dat, list(k=50, d=50, approximate=TRUE, ndist=10)))
```

Next make a new `SingleCellExperiment`  object with the combined data. In order 
to ensure that any cell can be traced first make a new sample identifier.

```{r combine-data}
for (i in 1:length(samples)){
  
  dat[[i]]$Sample_ID <- samples[i]
  
}

sce_colData <- Reduce(rbind, lapply(dat, function(x) colData(x)))
sce_rowData <- rowData(dat[[1]])

sce <- SingleCellExperiment(assays=list(counts=sce_counts),
                            colData=sce_colData,
                            rowData=sce_rowData)

#rm("sce_counts", "sce_rowData", "sce_colData")
#gc()

#sce_lib_factors <- c(unlist(sapply(dat, function(x) librarySizeFactors(x))))
#sce@int_colData$size_factor <- sce_lib_factors
sce <- normalize(sce)
reducedDim(sce, "MNN") <- dat_pca$corrected
```

Let's see whether this has improved things.

```{r plotting}
sce <- runPCA(sce)

plot_dat <- data.frame(pca1 = reducedDim(sce, "PCA")[,1],
                       pca2 = reducedDim(sce, "PCA")[,2],
                       mnn1 = reducedDim(sce, "MNN")[,1],
                       mnn2 = reducedDim(sce, "MNN")[,2],
                       sample = sce$Sample_ID)

ggplot(plot_dat, aes(x = pca1, y = pca2)) + 
  geom_point(aes(col = sample), alpha = 0.5) + 
  scale_colour_brewer(palette = "Set1")
ggplot(plot_dat, aes(x = mnn1, y = mnn2)) +  
  geom_point(aes(col = sample), alpha = 0.5) + 
  scale_colour_brewer(palette = "Set1")
```

## Clustering

### Dimension Reduction

```{r umap}
umap.defaults$n_neighbors = 20
umap.defaults$min_dist= 0.1
sce <- runUMAP(sce, use_dimred="MNN", config = umap.defaults)
```

### Clustering

```{r clustering}
snn.gr <- buildSNNGraph(sce, use.dimred="UMAP", k=100)
clusters <- igraph::cluster_louvain(snn.gr)
table(clusters$membership, sce$Sample_ID)
```

```{r plot-clustering}
sce$Cluster <- factor(clusters$membership)

plot_dat$umap1 <- reducedDim(sce, "UMAP")[,1]
plot_dat$umap2 <- reducedDim(sce, "UMAP")[,2]
plot_dat$cluster <- sce$Cluster 
plot_dat$umi <- sce$total_counts

ggplot(plot_dat, aes(x = umap1, y=umap2)) + 
  geom_point(aes(col = cluster), alpha = 0.5) + 
  scale_colour_brewer(palette = "Set1")
ggplot(plot_dat, aes(x = umap1, y=umap2)) + 
  geom_point(aes(col = sample), alpha = 0.5) + 
  scale_colour_brewer(palette = "Set1")
```

## Annotation

Highlight marker genes for different cell types.

```{r}
known_markers <- read.table(here::here("data/meta/", 
  "mouse_brain_cell_type_markers.txt"), sep="\t", fill=TRUE, header=T)

all_cell_types <- sapply(as.character(known_markers$Description), 
  function(x) strsplit(x, ",")[[1]][1])
known_markers$CellType <- all_cell_types

glia_markers <- sapply(which(known_markers$Neurotransmitter==""),
  function(x) unlist(strsplit(as.character(known_markers$Markers[x]), " ")))
vglut1_markers <- sapply(grep("VGLUT1", known_markers$Neurotransmitter),
  function(x) unlist(strsplit(as.character(known_markers$Markers[x]), " ")))
vglut2_markers <- sapply(grep("VGLUT2", known_markers$Neurotransmitter),
  function(x) unlist(strsplit(as.character(known_markers$Markers[x]), " ")))
GABA_markers <- sapply(grep("GABA", known_markers$Neurotransmitter),
  function(x) unlist(strsplit(as.character(known_markers$Markers[x]), " ")))


markers <- list(gaba = unlist(GABA_markers),
                glia = unlist(glia_markers),
                vglut1 = unlist(vglut1_markers),
                vglut2 = unlist(vglut2_markers))

cell_type_means <- sapply(markers, function(x) {
  index <- match(x , rowData(sce)$Symbol)
  index <- index[!is.na(index)]
  return(colMeans(as.matrix(logcounts(sce[index,]))))
})
colnames(cell_type_means) <- names(markers)

plot_dat <- cbind(plot_dat, cell_type_means)

for (i in colnames(cell_type_means)) {
  
  gg <- ggplot(plot_dat, aes(x = umap1, y = umap2)) + 
    geom_point(aes_string(color=i)) + scale_color_viridis_c()
  
  print(gg)
  
}

ggplot(plot_dat, aes(x = umap1, y = umap2)) + 
  geom_point(aes(color=umi)) + scale_color_viridis_c()
```

## Trajectory analysis

Still to come!
