---
title: 'snRNA-seq workflow: Seurat'
author: "Saskia Freytag"
date: "09/04/2019"
output:
  html_document:
    toc: true
    theme: simplex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
```

## Prerequisities 

### Required libraries

```{r libraries}
library(dplyr)
library(scater)
library(Seurat)
library(sctransform)
```

```{r setup-ggplot2}
library(ggplot2)
theme_set(theme_classic())
```

## Import data

### Loading data

```{r import}
samples <- list.files(here::here("data/raw"))
dat_s <- lapply(samples, function(x) Read10X(here::here("data/raw", x, "mm10")))
dat_s <- lapply(dat_s, function(x) CreateSeuratObject(x, 
    min.cells=(0.01*dim(x)[2])))
names(dat_s) <- samples
```

```{r}
for (i in 1:length(samples)){
  
  dat_s[[i]]$Sample_ID <- samples[i]
  
}
```

## Filtering

### Filtering of nuclei

First, we filter any nuclei with greater than 10\% total mitochondrial 
contamination. Fortunatley, there are no nuclei in this test dataset. But this 
is an importnat quality control step.

```{r filter-mito}
for (i in 1:length(dat_s)) {
  
  dat_s[[i]][["percent.mt"]] <- PercentageFeatureSet(object = dat_s[[i]], 
    pattern = "^mt-")

}

no_nuclei_mito <- sapply(dat_s, function(x) sum(x$percent.mt>10))

paste0("Nuclei with more than 10`% mitochondrial percentage in ", samples, ": ",
      no_nuclei_mito, " (", round(no_nuclei_mito/sapply(dat_s, 
        function(x) dim(x)[2]),2)*100, "%)")

dat_s <- lapply(dat_s, function(x) subset(x, subset = percent.mt<10))
```

Next we filter any nuclei that have library sizes three standard below the 
median. We also filter any nuclei with a number of features three standard
deviations below the median.

```{r filter-low}
libsize_drop <- lapply(dat_s, function(x) isOutlier(x$nCount_RNA, nmads=3,
  log=TRUE, type="lower"))
feature_drop <-  lapply(dat_s, function(x) isOutlier(x$nFeature_RNA, nmads=3,
  log=TRUE, type="lower"))

no_nuclei_libsize <- sapply(libsize_drop, function(x) sum(x))
no_nuclei_feature <- sapply(feature_drop, function(x) sum(x))

paste0("Nuclei not passing library size filter ", samples, ": ",
      no_nuclei_libsize, " (", round(no_nuclei_libsize/sapply(dat_s, function(x) dim(x)[2]),2)*100, "%)")

paste0("Nuclei not passing feature filter ", samples, ": ",
      no_nuclei_feature, " (", round(no_nuclei_feature/sapply(dat_s, function(x) dim(x)[2]),2)*100, "%)")

for (i in 1:length(samples)){
  
  dat_s[[i]] <- dat_s[[i]][, !(libsize_drop[[i]]|feature_drop[[i]])]

}
```

## Normalization

```{r normalization}
dat_s <- lapply(dat_s, function(x) NormalizeData(object = x, verbose = FALSE))
```

## Find variable genes

```{r variable-genes}
dat_s <- lapply(dat_s, function(x) FindVariableFeatures(object = x, 
  selection.method = "vst", nfeatures = 500))
```

### Batch correction

Perfom mnn correction to generate a common PCA representation for the samples.

```{r batch-correction}
seurat_anchors <- FindIntegrationAnchors(object.list = dat_s, dims = 1:20)
seurat <- IntegrateData(anchorset = seurat_anchors, dims = 1:20)
```

Next make a new `Seurat`  object with the combined data. In order 
to ensure that any cell can be traced first make a new sample identifier.

```{r combine-data}
DefaultAssay(object = seurat) <- "integrated"
seurat <- ScaleData(object = seurat, verbose = FALSE)

seurat <- RunPCA(object = seurat, npcs = 30, verbose = FALSE)
```

Let's see whether this has improved things.

```{r plotting}
DimPlot(object = seurat, reduction = "pca", group.by = "Sample_ID")
```

## Clustering

### Dimension Reduction

```{r umap}
seurat <- RunUMAP(object = seurat, reduction = "pca", dims = 1:20)
```

### Clustering

```{r clustering}
seurat <- FindNeighbors(object = seurat, reduction = "umap", 
  dims = 1:2, k.param = 50)
seurat <- FindClusters(seurat, resolution = 0.5)
```

```{r plot-clustering}
DimPlot(object = seurat, reduction = "umap", group.by = "seurat_clusters")
DimPlot(object = seurat, reduction = "umap", group.by = "Sample_ID")
```

## Annotation

Highlight marker genes for different cell types.

```{r plotting}
plot_dat <- data.frame(umap1 = seurat@reductions$umap@cell.embeddings[,1],
                       umap2 = seurat@reductions$umap@cell.embeddings[,2],
                       sample = seurat$Sample_ID,
                       cluster = seurat$seurat_clusters,
                       umi = seurat@meta.data$nCount_RNA)

ggplot(plot_dat, aes(x = umap1, y = umap2)) + 
  geom_point(aes(col = sample), alpha = 0.5) + 
  scale_colour_brewer(palette = "Set1")
ggplot(plot_dat, aes(x = umap1, y = umap2)) +  
  geom_point(aes(col = cluster), alpha = 0.5) + 
  scale_colour_brewer(palette = "Set1")
```


```{r}
known_markers <- read.table(here::here("data/meta/", 
  "mouse_brain_cell_type_markers.txt"), sep="\t", fill=TRUE, header=T)

all_cell_types <- sapply(as.character(known_markers$Description), 
  function(x) strsplit(x, ",")[[1]][1])
known_markers$CellType <- all_cell_types

glia_markers <- sapply(which(known_markers$Neurotransmitter==""),
  function(x) unlist(strsplit(as.character(known_markers$Markers[x]), " ")))
vglut1_markers <- sapply(grep("VGLUT1", known_markers$Neurotransmitter),
  function(x) unlist(strsplit(as.character(known_markers$Markers[x]), " ")))
vglut2_markers <- sapply(grep("VGLUT2", known_markers$Neurotransmitter),
  function(x) unlist(strsplit(as.character(known_markers$Markers[x]), " ")))
GABA_markers <- sapply(grep("GABA", known_markers$Neurotransmitter),
  function(x) unlist(strsplit(as.character(known_markers$Markers[x]), " ")))


markers <- list(gaba = unlist(GABA_markers),
                glia = unlist(glia_markers),
                vglut1 = unlist(vglut1_markers),
                vglut2 = unlist(vglut2_markers))

cell_type_means <- sapply(markers, function(x) {
  index <- match(x , rownames(seurat@assays$RNA))
  index <- index[!is.na(index)]
  return(colMeans(as.matrix(seurat@assays$RNA[index,])))
})
colnames(cell_type_means) <- names(markers)

plot_dat <- cbind(plot_dat, cell_type_means)

for (i in colnames(cell_type_means)) {
  
  gg <- ggplot(plot_dat, aes(x = umap1, y = umap2)) + 
    geom_point(aes_string(color=i)) + scale_color_viridis_c()
  
  print(gg)
  
}

ggplot(plot_dat, aes(x = umap1, y = umap2)) + 
  geom_point(aes(color=umi)) + scale_color_viridis_c()
```

## Investigate mature genes

```{r}
mature_genes <- c("Map2", "Rbfox3")
index <- match(mature_genes , rownames(seurat@assays$RNA))
index <- index[!is.na(index)]
plot_dat$mature <- colMeans(as.matrix(seurat@assays$RNA[index,]))

ggplot(plot_dat, aes(x = umap1, y = umap2)) + 
  geom_point(aes(color=mature)) + scale_color_viridis_c()
```
