---
title: 'snRNA-seq workflow: bioconductor'
author: "Saskia Freytag"
date: "08/04/2019"
output:
  html_document:
    toc: true
    theme: simplex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
```

## Prerequisities 

### Required libraries

```{r libraries}
library(DropletUtils)
library(scater)
library(scran)
library(batchelor)
library(umap)
```

```{r setup-ggplot2}
library(iSEE)
library(ggplot2)
theme_set(theme_classic())
```

## Import data

### Loading data

Unfortunately, Cell Ranger recently changed its naming convention. The file that
used to be called `genes.tsv` is now named `features.tsv`. The easiest 
work-around, until a change to `DropUtils`, is to rename the files.

```{r import}
samples <- list.files(here::here("data/raw"))
dat <- lapply(samples, function(x)
  read10xCounts(here::here("data/raw", x, "mm10")))
names(dat) <- samples
```

### Combine data

```{r combine-data}
for (i in 1:length(samples)) {
  dat[[i]]$Sample_ID <- samples[i]
}

sce <- do.call(cbind, dat)
sce$Sample_ID <- as.factor(sce$Sample_ID)
```

## Filtering

### Filtering of nuclei

First, we filter any nuclei with greater than 10\% total mitochondrial 
contamination. 

```{r filter-mito}
rowData(sce)$Mito <- grepl("^mt-", rowData(sce)$Symbol)
sce <- calculateQCMetrics(sce,
  feature_controls = list(Mt = rowData(sce)$Mito)
)

no_nuclei_mito <- sapply(levels(sce$Sample_ID), function(x)
  sum((colData(sce)$pct_counts_Mt > 10)[sce$Sample_ID == x]))

paste0(
  "Nuclei with more than 10% mitochondrial percentage in ", samples, ": ",
  no_nuclei_mito, " (", round(no_nuclei_mito / sapply(
    levels(sce$Sample_ID),
    function(x) sum(sce$Sample_ID == x)
  ), 2) * 100, "%)"
)

sce <- sce[, !colData(sce)$pct_counts_Mt > 10]
```

Next we filter any nuclei that have library sizes three standard below the 
median. We also filter any nuclei with a number of features three standard
deaviations below the median.

```{r filter-low}
libsize_drop <- isOutlier(sce$total_counts,
  nmads = 3,type = "lower", log = TRUE
)
feature_drop <- isOutlier(sce$total_features_by_counts,
  nmads = 3, type = "lower", log = TRUE
)

no_nuclei_libsize <- sapply(levels(sce$Sample_ID), function(x)
  sum(libsize_drop[sce$Sample_ID == x]))
no_nuclei_feature <- sapply(levels(sce$Sample_ID), function(x)
  sum(feature_drop[sce$Sample_ID == x]))

paste0(
  "Nuclei not passing library size filter ", samples, ": ",
  no_nuclei_libsize, " (", round(no_nuclei_libsize /
    sapply(
      levels(sce$Sample_ID),
      function(x) sum(sce$Sample_ID == x)
    ), 2) * 100, "%)"
)

paste0(
  "Nuclei not passing feature filter ", samples, ": ",
  no_nuclei_feature, " (", round(no_nuclei_feature /
    sapply(
      levels(sce$Sample_ID),
      function(x) sum(sce$Sample_ID == x)
    ), 2) * 100, "%)"
)

sce <- sce[, !(libsize_drop | feature_drop)]
```

### Filtering of genes

## Filter mitochondrial genes

```{r filter-mito-genes}
sce <- sce[!rowData(sce)$Mito, ]
```

## Filter genes by expression

In order to filter genes we need to combine the data. Then we filter any gene
that is seen in less than 1\% of the data.

```{r filter-genes}
keep_genes <- apply(counts(sce), 1, function(x)
  sum(x > 0) > (dim(sce)[2] * 0.01))

sce <- sce[keep_genes, ]
```

## Normalization

```{r normalization}
sce <- normalize(sce)
```

### Batch correction

Perfom mnn correction to generate a common PCA representation for the samples.

```{r batch-correction}
dat_pca <- batchelor::fastMNN(sce,
  batch = sce$Sample_ID,
  k = 50, d = 50, ndist = 10
)

reducedDim(sce, "corrected") <- reducedDim(dat_pca, "corrected")
```

Next make a new `SingleCellExperiment`  object with the combined data. In order 
to ensure that any cell can be traced first make a new sample identifier.

Let's see whether this has improved things.

```{r plotting}
sce <- runPCA(sce)

plot_dat <- data.frame(
  pca1 = reducedDim(sce, "PCA")[, 1],
  pca2 = reducedDim(sce, "PCA")[, 2],
  mnn1 = reducedDim(sce, "corrected")[, 1],
  mnn2 = reducedDim(sce, "corrected")[, 2],
  sample = sce$Sample_ID
)

ggplot(plot_dat, aes(x = pca1, y = pca2)) +
  geom_point(aes(col = sample), alpha = 0.5) +
  scale_colour_brewer(palette = "Set1")
ggplot(plot_dat, aes(x = mnn1, y = mnn2)) +
  geom_point(aes(col = sample), alpha = 0.5) +
  scale_colour_brewer(palette = "Set1")
```

## Clustering

### Dimension Reduction

```{r umap}
sce <- runUMAP(sce, use_dimred = "corrected", n_neighbors = 20)
```

### Clustering

```{r clustering}
snn.gr <- buildSNNGraph(sce, use.dimred = "UMAP", k = 100)
clusters <- igraph::cluster_louvain(snn.gr)
table(clusters$membership, sce$Sample_ID)
```

```{r plot-clustering}
sce$Cluster <- factor(clusters$membership)

plot_dat$umap1 <- reducedDim(sce, "UMAP")[, 1]
plot_dat$umap2 <- reducedDim(sce, "UMAP")[, 2]
plot_dat$cluster <- sce$Cluster
plot_dat$umi <- sce$total_counts

ggplot(plot_dat, aes(x = umap1, y = umap2)) +
  geom_point(aes(col = cluster), alpha = 0.5) +
  scale_colour_brewer(palette = "Set1")
ggplot(plot_dat, aes(x = umap1, y = umap2)) +
  geom_point(aes(col = sample), alpha = 0.5) +
  scale_colour_brewer(palette = "Set1")
```

## Annotation

Highlight marker genes for different cell types.

```{r}
known_markers <- read.table(here::here(
  "data/meta/",
  "mouse_brain_cell_type_markers.txt"
), sep = "\t", fill = TRUE, header = T)

all_cell_types <- sapply(
  as.character(known_markers$Description),
  function(x) strsplit(x, ",")[[1]][1]
)
known_markers$CellType <- all_cell_types

glia_markers <- sapply(
  which(known_markers$Neurotransmitter == ""),
  function(x) unlist(strsplit(as.character(known_markers$Markers[x]), " "))
)
vglut1_markers <- sapply(
  grep("VGLUT1", known_markers$Neurotransmitter),
  function(x) unlist(strsplit(as.character(known_markers$Markers[x]), " "))
)
vglut2_markers <- sapply(
  grep("VGLUT2", known_markers$Neurotransmitter),
  function(x) unlist(strsplit(as.character(known_markers$Markers[x]), " "))
)
GABA_markers <- sapply(
  grep("GABA", known_markers$Neurotransmitter),
  function(x) unlist(strsplit(as.character(known_markers$Markers[x]), " "))
)


markers <- list(
  gaba = unlist(GABA_markers),
  glia = unlist(glia_markers),
  vglut1 = unlist(vglut1_markers),
  vglut2 = unlist(vglut2_markers)
)

cell_type_means <- sapply(markers, function(x) {
  index <- match(x, rowData(sce)$Symbol)
  index <- index[!is.na(index)]
  return(colMeans(as.matrix(logcounts(sce[index, ]))))
})
colnames(cell_type_means) <- names(markers)

plot_dat <- cbind(plot_dat, cell_type_means)

for (i in colnames(cell_type_means)) {
  gg <- ggplot(plot_dat, aes(x = umap1, y = umap2)) +
    geom_point(aes_string(color = i)) + scale_color_viridis_c()

  print(gg)
}

ggplot(plot_dat, aes(x = umap1, y = umap2)) +
  geom_point(aes(color = umi)) + scale_color_viridis_c()
```

## Investigate mature genes

```{r}
mature_genes <- c("Map2", "Rbfox3")
index <- match(mature_genes, rowData(sce)$Symbol)
index <- index[!is.na(index)]
plot_dat$mature <- colMeans(as.matrix(logcounts(sce[index, ])))

ggplot(plot_dat, aes(x = umap1, y = umap2)) +
  geom_point(aes(color = mature)) + scale_color_viridis_c()
```

## Rescaling of batches

```{r rescaling}
sce_rescale <- rescaleBatches(sce, batch = sce$Sample_ID)
```

## Trajectory analysis

Still to come!

## Interactive instance

### iSEE

```{r iSEE}   
rownames(sce) <- rowData(sce)$Symbol
sce <- sce [!duplicated(rownames(sce)), ]

fexArg <- featAssayPlotDefaults(sce, 2)
fexArg$YAxisFeatName <- c("Map2", "Rbfox3")
fexArg$ColorBy <- c("Column data", "Column data")
fexArg$ColorByColData <- c("Cluster", "Sample_ID")

rdArgs <- redDimPlotDefaults(sce, 2)
rdArgs$DataBoxOpen <- FALSE
rdArgs$Type <- c("UMAP", "corrected")
rdArgs$ColorBy <- c("Column data", "Column data")
rdArgs$ColorByColData <- c("Cluster", "Sample_ID")
rdArgs$XAxis <- c(1, 1)
rdArgs$YAxis <- c(2, 2)

init <- DataFrame(
  Name = c(
    "Reduced dimension plot 1", "Feature assay plot 1",
    "Reduced dimension plot 2", "Feature assay plot 2"
  ),
  Width = c(6, 6)
)

app <- iSEE(sce,
  initialPanels = init, featAssayArgs = fexArg,
  redDimArgs = rdArgs
)
```

To initialize simply run `app`.
